The importance of transient resonances in extreme-mass-ratio inspirals (new title?)

### 2015-06-29 res_char.tex
Some tweaks

Flanagan and Hinderer~\cite{Flanagan2012} highlighted a previously overlooked phenomenon that occurs in the general relativistic two-body problem, that of transient resonances. Geodesic orbits in GR have three associated frequencies: the radial frequency $\Omega_r$, the polar frequency $\Omega_\theta$ and the azimuthal frequency $\Omega_\phi$. The first two describe libration and the third rotation (except in the case of polar orbits where $\Omega_\theta$ also describes rotation)~\cite{Goldstein2002}. % Section 10.6
In the weak-field limit, these all tend towards the Keplerian frequency; in the strong-field, $\Omega_r < \Omega_\theta < \Omega_\phi$, and they may differ significantly. For extreme-mass-ratio systems, the evolution time-scale is much longer than the orbital period such that the motion of the smaller body is approximately geodesic over orbital time-scales. The evolution of the orbit can be approximated as a series of geodesics using the osculating element formalism~\cite{Pound2008,Gair2011a}. During this evolution, the frequencies may become commensurate. Resonances occur when the radial and polar frequencies are rational multiples of each other:
\begin{equation}
\nu = \frac{\Omega_r}{\Omega_\theta} = \frac{n_\theta}{n_r},
\end{equation}
where $n_r$ and $n_\theta$ are integers (with no common factors). During resonance, terms in the self-force that usually average to zero can combine coherently, significantly impacting the orbital motion~\cite{Flanagan2012a}.

@@

In \secref{problem}, we formulate the specific problem at hand: that of geodesic motion in Kerr spacetime, perturbed by the gravitational self-force. We then study generic properties of transient resonances in \secref{properties}, detailing their location in parameter space, the time-scales over which they affect the motion and the resulting GW flux enhancements. Specific examples are considered to illustrate the effects of resonances at the start of \secref{astrophysics}, before finally turning to an astrophysical population in \secref{population}. Our conclusions can be found in \secref{conclusion}.

@@

To construct an adiabatic trajectory we need the $2$-torus-averaged fluxes of our osculating elements. Computing an average of a quantity over the $\{q_r, q_\theta\}$ is trivial if it is parametrized in terms of these variables,
\begin{equation}
\left\langle \diff{X}{\lambda}\right\rangle_{q_r,\,q_\theta} = \frac{1}{(2\pi)^2}\intd{0}{2\pi}{\intd{0}{2\pi}{\diff{X}{\lambda}}{q_r}}{q_\theta}.
\end{equation}
However, we are using $\psi$ and $\chi$, as these are simpler to evolve; furthermore, we compute instantaneous co-ordinate time fluxes $\dot{X}$, not Mino-time fluxes. Changing variables gives an average of \cite{Drasco2005}
\begin{widetext}\begin{align}
\left\langle \diff{X}{\lambda}\right\rangle_{q_r,\,q_\theta} = {} & \frac{1}{\Lambda_r \Lambda_\theta}\intd{0}{2\pi}{\intd{0}{2\pi}{\left(\diff{\psi}{t}\right)^{-1} \left(\diff{\chi}{t}\right)^{-1} \left(\diff{t}{\lambda}\right)^{-2} \diff{X}{\lambda}}{\psi}}{\chi} \\
 = {} & \frac{1}{\Lambda_r \Lambda_\theta}\intd{0}{2\pi}{\intd{0}{2\pi}{\left(\diff{\psi}{t}\right)^{-1} \left(\diff{\chi}{t}\right)^{-1} \left(\diff{t}{\lambda}\right)^{-1} \dot{X}}{\psi}}{\chi}.
\label{eq:2torus-average}
\end{align}\end{widetext}
This average describes the Mino-time rate of change of the quantity $X$ over an orbit. To convert to a co-ordinate flux of the averaged quantity, we simply divide by the period $\Gamma$, defining
\begin{equation}
\dot{\left\langle X\right\rangle}_{q_r,\,q_\theta} = \frac{1}{\Gamma}\left\langle \diff{X}{\lambda}\right\rangle_{q_r,\,q_\theta}.
\end{equation}
Is in convenient to calculate $\Gamma$ as
\begin{equation}
\Gamma = \left\langle \diff{t}{\lambda}\right\rangle_{q_r,\,q_\theta},
\end{equation}
using \eqnref{2torus-average}, as this allows us to eliminate $\Lambda_r$ and $\Lambda_\theta$ from the calculation. We compute these integrals using a $300\times 300$ grid of $\{\psi,\chi\}$ values and employing a Newton--Cotes approximation in each dimension. The procedure requires $\mathcal{O}(10^5)$ separate evaluations of the derivatives at each time-step of an evolution, and so is computationally expensive to perform. The adiabatic derivatives vary on much longer timescales than the orbital motion, and so in practice, we can make use of interpolation. We calculate the adiabatic derivatives, as above, at three points in the inspiral spaced by $\dd t/\eta$ where $\dd t$ is the orbital time-step. When the system is close to the central point, we use a quadratic fit to estimate the adiabatic derivatives. Once the system evolves close to one of the boundaries, we update the interpolation points, taking the closest boundary to be the new central point. The averaged fluxes successfully describe the leading-order secular evolution of the trajectory, as illustrated in \figref{good-traj}.

@@

The first step in studying the effect of transient resonances is to locate orbital parameters for which the frequencies are commensurate. We can calculate the frequencies and so we are left with the problem of solving $\Omega = n_r \Omega_r - n_\theta \Omega_\theta = 0$ numerically. When considering the full parameter set of $\{p,e,\iota,a_\ast,\nu\}$, it is apparent that the search for resonances becomes expensive as a consequence of the dimensionality. It is therefore useful to have a guide of where to look. In \apref{location} we build a simple approximate model as a starting point for the numerical search. The resonances occur at relatively small periapses, corresponding to regions of strong-field gravity. Having located where in an inspiral we can expect to encounter a transient resonance, we must now consider its impact. In \secref{res-time} we determine the characteristic time-scales describing resonance, and in \secref{flux-enhance} we look at the impact of passing through a resonance on the evolution of the orbit.

@@

We can study the predicted orbital parameter jumps numerically by considering the long-time-averaged fluxes for particular (fixed) geodesics. For non-resonant geodesics, this will depend only on the orbital shape parameters, but for resonant orbits there will be an additional dependence on the initial phase difference. We consider a selection of different orbital shape parameters corresponding to low-order transient resonances and calculate the time-averaged fluxes for $50$ different initial phases. We evolve for an integer number of cycles of both the radial and polar motions. Specifically, when considering a $n_\theta$:$n_r$ resonance, we take $T = 2\pi N n_\theta/\Omega_r = 2\pi N n_r/\Omega_\theta$ for some integer $N$. We have found that taking $N=5$ gives sufficiently accurate results. We can then estimate the jumps by taking the difference between the maximum and minimum time-averaged fluxes.

@@

It is informative to consider the dependence on the resonance ratio $\nu$. Fixing the eccentricity at $e=0.95$ to emphasise the variation, we compute the flux enhancements for different resonance ratios, shown in \figref{fig:res-flux-rms-p-vsnu}, where we also highlight ratios arising from different values of $n_\theta$. We observe a clear relationship between the flux enhancement and $\nu$ for all resonances with $n_\theta = 1$ or $2$, independent of which of those is considered. However, negligibly small flux enhancements are observed for systems with $n_\theta = 3$ or higher. This can be explained by the quadrupolar nature of GWs. A resonant trajectory averaged over time (and consequently $\phi$) takes $n_\theta$ (possibly repeated) values of $\theta$ at each radial turning point. Considering rings of matter at these coordinates, the dominant spherical harmonic describing the resultant mass distribution has $l=n_\theta$ (and $m=0$). Since GWs are generated by the quadrupole moment at leading order, the most significant resonant enhancements will arise from systems with $n_\theta = 2$. Due to the presence of higher harmonics, $n_\theta = 1$ systems also contribute strongly; it is perhaps more appropriate to treat $1$:$x$ resonances as \emph{de facto} $2$:$2x$ resonances, explaining the linked behaviour of the flux enhancements for $n_\theta = 1$ and $2$ systems.

@@

The resonant enhancements to the fluxes of $E$, $L_z$ and $Q$ have been calculated previously using Teukolsky-based methods for a handful of systems~\cite{Flanagan2012a}. They consider the 1:2, 1:3, 2:3 and 3:4 resonances for EMRIs with $a=0.9$, either $e=0.3$ or $0.7$, and $\theta_\mathrm{inc} = 20\text{\textdegree}$ or $70$\textdegree. We compare our results to their published values (in their tables I--IV) in \figref{res-flux-FHR}. There is some qualitative agreement between the two approaches: more eccentric orbits give rise to larger enhancements; and the 3:4 resonance has a negligibly small enhancement. However, for the most significant 2:3 resonance, we typically predict larger orbital parameter jumps. Our results in \secref{astrophysics}, regarding the impact of resonances on a population of EMRIs, are therefore conservative predictions, as our model over-estimates the resonance effects.

\begin{figure*}[htbp]
\centering
\includegraphics[width=\textwidth]{res_flux_FHR}
\caption{\label{fig:res-flux-FHR}Relative flux enhancements of $E$ (left column), $L_z$ (middle) and $Q$ (right) for resonance ratios corresponding to the 1:3, 1:2, 2:3 and 3:4 resonances. We show results as computed here with a solid red line, and results from~\cite[FHR14]{Flanagan2012a} with a dashed blue line. Each row has a different eccentricity and inclination corresponding to tables I--IV in FHR14. All systems have $a=0.9$.}
\end{figure*}

@@

\subsection{Waveform model}
\emph{Can probably shorten this section a bit.}

To generate gravitational waveforms, we choose to employ the numerical kludge (NK) method of Babak {\it{et al}}.~\cite{Babak2007}, augmented to include evolution of the positional elements. The principle behind the class of kludge methods is to first compute inspiral trajectories and then separately (and not necessarily consistently) calculate the GW emission sourced by a compact object moving along that trajectory. This can be much quicker and easier to calculate than waveforms using Teukolsky-based methods (currently the most accurate prescription available) and yet gives similar results; overlaps between Teukolsky-based and the best NK waveforms are typically 95\% or higher for a variety of orbits~\cite{Babak2007, Berry2013}.

Following the NK method, we first compute the inspiral trajectory of the compact object around a central Kerr BH using the method of osculating elements detailed in \secref{forced-motion}, with the dissipative part of the post-Newtonian self-force from \secref{self-force}. We then interpret the Boyer--Lindquist coordinates as spherical polar coordinates in Minkowski, facilitating the use of a flat-spacetime waveform generation technique; specifically, we apply the standard quadrupole formula~\cite{Misner1973}.

We expect these waveforms to be sufficiently accurate for our purposes, given the inaccuracy of the self-force model being employed. It would, however, be straight-forward to adopt a different waveform generation scheme, for example by including higher order multipole moments, should advances be made in determining the self-force.

Once we have established a waveform generation procedure, we can define a natural inner product between two waveforms, $s(t)$ and $h(t)$, according to
\begin{equation}
\label{eq:innerprod}
\overlap{s}{h} = 2 \intd{0}{\infty}{\frac{\tilde{s}(f)\tilde{h}^*(f)+\tilde{s}^*(f)\tilde{h}(f)}{S_n(f)}}{f},
\end{equation}
where $\tilde{s}(f)$ represents the Fourier transform of $s(t)$ and similarly for $\tilde{h}(f)$, and $S_n(f)$ is the one-sided noise power spectral density (PSD). We are interested in the detectability of EMRIs using future space-based GW detectors and so we use a PSD appropriate for describing the eLISA mission concept. We use the analytic approximation of Amaro-Seoane {\it{et al}}.~\cite{Amaro-Seoane2012a}

In practice, the continuous functions $s(t)$ and $h(t)$ are not available, instead replaced by the time series $s(t_i)$ and $h(t_j)$ where $\{t_i\}$ and $\{t_j\}$ are equally-spaced sets of time values, but which do not necessarily have the same length or spacing. To take this into account, when evaluating overlaps, we first project the $h(t_j)$ onto $\{t_i\}$ using a cubic interpolation scheme, setting any values outside the domain of $\{t_j\}$ to zero. A Planck-taper window function~\cite{McKechan2010} is applied to the resulting time series to reduce unwanted spectral leakage in the Fourier transforms.

After tapering the projected waveforms, we then calculate overlaps using \eqnref{innerprod} but with discrete Fourier transforms and the integral over frequency replaced with an appropriate sum over Fourier modes. In addition, we maximise every overlap calculation with respect to some time-offset between the models by computing the correlation between the waveforms. This can be performed cheaply using Fourier transforms and can provide a significant increase to the overlap, even though typical offset times are shorter than the orbital timescale.

@@

We wish to test whether there exists some set of parameters $\Theta^a$ such that the resulting adiabatic waveform is sufficiently similar to the full waveform. We do this by evaluating the signal-to-noise ratio (SNR) for each waveform template, defined by
\begin{equation}
\rho\left[h\right] = \frac{\overlap{s}{h}}{\sqrt{\overlap{h}{h}}}.
\end{equation}
If a template were to exactly match the data, it would produce an SNR of $\sqrt{\overlap{s}{s}}$ and so the overlap
\begin{equation}
\label{eq:overlap}
\mathbb{O}\left[h\right] = \frac{\overlap{s}{h}}{\sqrt{\overlap{s}{s}}\sqrt{\overlap{h}{h}}}
\end{equation}
provides an indication of how well-matched the template is to the data. We leave an investigation into parameter estimation biases (where the parameters of the best-matching adiabatic $\Theta^a$ do not agree with the parameters of the system $\bar{\Theta}^a$) to future work.

To demonstrate the approach, we calculate the full evolution and the 2-torus averaged adiabatic evolution over $2$ years for a set of initial parameters chosen to avoid any significant resonances. Specifically, we choose: the mass of the CO $\mu = 10 M_{\odot}$; the mass ratio $\mu / M = 3\times 10^{-6}$ where $M$ is the mass of the central BH; the spin of the central BH $a=0.95$; the initial dimensionless semi-latus rectum $p=7.5$; the initial eccentricity $e=0.7$; the initial inclination angle $\cos \iota = 0.5$; and the redshift of the source $z=0.204$. The remaining intrinsic (the initial phases $\psi$ and $\chi$) and extrinsic (the position of the source on the sky, the direction of the spin of the central BH and the azimuthal angle) parameters are set to arbitrary values. The resulting evolutions of the orbital parameters $E$, $L_z$ and $Q$ are shown in \figref{good-traj}, where it can be seen that the adiabatic evolution closely matches the full evolution on longer inspiral timescales. The inset plots show the start of the evolution, on a timescale associated with the orbital motion of the CO; the 2-torus averaging explicitly smoothes out the visible structure on this scale.

\begin{figure*}[htbp]
\centering
\includegraphics[width=\textwidth]{Fig_good_traj}
\caption{\label{fig:good-traj}The evolution of the orbital parameters $E$ (left), $L_z$ (center) and $Q$ (right) under the full (solid line) and adiabatic (dashed) models for an illustrative EMRI system that does not encounter any significant resonances. The inset plots show the behaviour on short timescales, where the fast orbital oscillations can be seen.}
\end{figure*}

@@

\begin{figure*}[htbp]
\centering
\includegraphics[width=0.92\textwidth]{Fig_good_waveform}
\caption{\label{fig:good-waveform}The plus-polarised waveform for an illustrative EMRI system with $p=7.5$, shown for two short segments at the start and end of a $2$ year evolution. The full and adiabatic models are both shown, but are indistinguishable by eye.}
\end{figure*}

@@

We generate a representative EMRI population by making reasonable assumptions about the distributions of the 14 parameters in the analytic kludge model of Barack and Cutler \cite{Barack2004}. The EMRI event rate is dominated by the inspiral of heavier BHs due to the larger observational volume in comparison to that for white dwarfs and neutron stars~\cite{gair_event_2004}. We therefore take the mass of the compact object to be $\mu = 10 M_\odot$, corresponding to a typical mass for stellar BHs.

@@

Despite the difficulties, simple estimates of $\mathcal{R}$ have been carried out using Monte Carlo methods to count the number of stars from isothermal distributions that spiral in to a SMBH without plunging. The result is a scaling law for each species of compact object of the form
\begin{equation}
\label{eq:EMRI-intrinsic-rate}
\mathcal{R}(M) = \mathcal{R}_0 \left(\frac{M}{M_0}\right)^\alpha,
\end{equation}
where $M$ is the mass of the SMBH and $M_0 = 3\times 10^6 M_\odot$ is some fiducial mass. \Citet{hopman_extreme_2009} finds that $\alpha = \{-0.15,\,-0.25,\,-0.25\}$ for BHs, NSs and WDs, with respective event rates $\mathcal{R}_0 = \{400,\, 7,\, 20\}\, \mathrm{Gyr}^{-1}$ for each component. This neglects the effects of resonant relaxation, ignores the spin of the SMBH, and assumes that the $M$--$\sigma$ relation holds for all SMBH masses. Each of these is likely to significantly impact the event rate, but \eqnref{EMRI-intrinsic-rate} can still be used as a rough guide to the expected number of events.

\Citet{amaro-seoane_impact_2011} studied the effects of mass segregation on the intrinsic EMRI event rate, using direct-summation $N$-body simulations to calibrate a Fokker-Planck description for the bulk properties of the stellar distribution. They found a better fit for the power law spectral index for BHs of $\alpha = -0.19$.

The comoving number density of galaxies is challenging to estimate due to local structure in the Universe, the evolution of that structure, and properties of the SMBHs themselves. We simplify the problem by assuming a homogeneous distribution that does not evolve with redshift, which is reasonable for the typical scales considered by GW detectors. We also neglect correlations between the SMBH mass and spin, and impose a power law scaling relation for the comoving number density
\begin{equation}
\diff{n}{\ln M}(M) = n_0 \left(\frac{M}{M_0}\right)^\beta.
\end{equation}
There is significant uncertainty in the SMBH mass function, but this simple functional form is found to be in good agreement with observations from the Sloan Digital Sky Survey for the mass range of interest; it is sufficient to set $\beta = 0$ and $n_0 = 0.002 \, \mathrm{Mpc}^{-3}$ for SMBHs with $M < \mathcal{O}(10^7 M_\odot)$ \citep{greene_mass_2007, gair_lisa_2010}.

>>>>

Flanagan and Hinderer~\cite{Flanagan2012} highlighted a previously overlooked phenomenon that occurs in the general relativistic two-body problem, that of transient resonances. Geodesic orbits in GR have three associated frequencies: the radial frequency $\Omega_r$, the polar frequency $\Omega_\theta$ and the azimuthal frequency $\Omega_\phi$. The first two describe libration and the third rotation (except in the case of polar orbits where $\Omega_\theta$ also describes rotation)~\cite{Goldstein2002}. % Section 10.6
In the weak-field limit, these all tend towards the Keplerian frequency; in the strong-field, $\Omega_r < \Omega_\theta < \Omega_\phi$, and they may differ significantly. For extreme-mass-ratio systems, the evolution time-scale is much longer than the orbital period such that the motion of the smaller body is approximately geodesic over orbital time-scales. The evolution of the orbit can be approximated as a series of geodesics using the osculating element formalism~\cite{Pound2008,Gair2011a}. During this evolution, the frequencies may become commensurate. Resonances occur when the radial and polar frequencies are rational multiples of each other:
\begin{equation}
\nu \equiv \frac{\Omega_r}{\Omega_\theta} = \frac{n_\theta}{n_r},
\end{equation}
where $n_r$ and $n_\theta$ are integers (with no common factors). During resonance, terms in the self-force that usually average to zero can combine coherently, significantly impacting the orbital motion~\cite{Flanagan2012a}.

@@

In \secref{problem}, we formulate the specific problem: that of geodesic motion in Kerr spacetime, perturbed by the gravitational self-force. We then study generic properties of transient resonances in \secref{properties}, detailing their location in parameter space, the time-scales over which they affect the motion and the resulting GW flux enhancements. Specific examples are considered to illustrate the effects of resonances at the start of \secref{astrophysics}, before finally turning to an astrophysical population in \secref{population}. Our conclusions can be found in \secref{conclusion}.

@@

To construct an adiabatic trajectory we need the $2$-torus-averaged fluxes of our osculating elements. Computing an average of a quantity over the $\{q_r, q_\theta\}$ is trivial if it is parametrized in terms of these variables,
\begin{equation}
\left\langle \diff{X}{\lambda}\right\rangle_{q_r,\,q_\theta} = \frac{1}{(2\pi)^2}\intd{0}{2\pi}{\intd{0}{2\pi}{\diff{X}{\lambda}}{q_r}}{q_\theta}.
\end{equation}
However, we are using $\psi$ and $\chi$, as these are simpler to evolve; furthermore, we compute instantaneous co-ordinate time fluxes $\dot{X}$, not Mino-time fluxes. Changing variables gives an average of \cite{Drasco2005}
\begin{widetext}\begin{align}
\left\langle \diff{X}{\lambda}\right\rangle_{q_r,\,q_\theta} = {} & \frac{1}{\Lambda_r \Lambda_\theta}\intd{0}{2\pi}{\intd{0}{2\pi}{\left(\diff{\psi}{t}\right)^{-1} \left(\diff{\chi}{t}\right)^{-1} \left(\diff{t}{\lambda}\right)^{-2} \diff{X}{\lambda}}{\psi}}{\chi} \\
 = {} & \frac{1}{\Lambda_r \Lambda_\theta}\intd{0}{2\pi}{\intd{0}{2\pi}{\left(\diff{\psi}{t}\right)^{-1} \left(\diff{\chi}{t}\right)^{-1} \left(\diff{t}{\lambda}\right)^{-1} \dot{X}}{\psi}}{\chi}.
\label{eq:2torus-average}
\end{align}\end{widetext}
This average describes the Mino-time rate of change of the quantity $X$ over an orbit. To convert to a co-ordinate flux of the averaged quantity, we simply divide by the period $\Gamma$, defining
\begin{equation}
\dot{\left\langle X\right\rangle}_{q_r,\,q_\theta} = \frac{1}{\Gamma}\left\langle \diff{X}{\lambda}\right\rangle_{q_r,\,q_\theta}.
\end{equation}
It is convenient to calculate $\Gamma$ as
\begin{equation}
\Gamma = \left\langle \diff{t}{\lambda}\right\rangle_{q_r,\,q_\theta},
\end{equation}
using \eqnref{2torus-average}, as this allows us to eliminate $\Lambda_r$ and $\Lambda_\theta$ from the calculation.\footnote{We compute these integrals using a $300\times 300$ grid of $\{\psi,\chi\}$ values and employing a Newton--Cotes approximation in each dimension. The procedure requires $\mathcal{O}(10^5)$ separate evaluations of the derivatives at each time-step of an evolution, and so is computationally expensive to perform. However, the adiabatic derivatives vary on much longer timescales than the orbital motion (see \secref{res-time}), and so in practice, we can interpolate.} %We calculate the adiabatic derivatives, as above, at three points in the inspiral spaced by $\delta t/\eta$ where $\delta t$ is the orbital time-step. When the system is close to the central point, we use a quadratic fit to estimate the adiabatic derivatives. Once the system evolves close to one of the boundaries, we update the interpolation points, taking the closest boundary to be the new central point.
The averaged fluxes successfully describe the leading-order secular evolution of the trajectory (as illustrated in \figref{good-traj}).

@@

The first step in studying the effect of transient resonances is to locate orbital parameters for which the frequencies are commensurate. We can calculate the frequencies and so we are left with the problem of solving $\Omega = n_r \Omega_r - n_\theta \Omega_\theta = 0$ numerically. When considering the full parameter set of $\{p,e,\iota,a_\ast,\nu\}$, it is apparent that the search for resonances becomes expensive as a consequence of the dimensionality. It is therefore useful to have a guide of where to look. In \apref{location} we build a simple approximate model as a starting point for the numerical search. The resonances occur at relatively small periapses, corresponding to regions of strong-field gravity. Having located where in an inspiral we can expect to encounter a transient resonance, we must now consider its impact. In \secref{res-time} we determine the characteristic time-scales describing resonance, and in \secref{flux-enhance} we look at the impact of passing through a resonance on the evolution of the orbit, including some numerical results.

@@

We can study the predicted orbital-parameter jumps numerically by considering the long-time-average fluxes for particular (fixed) geodesics. For non-resonant geodesics, this will depend only on the orbital shape parameters, but for resonant orbits there will be an additional dependence on the initial phase difference. We consider a selection of different orbital shape parameters corresponding to low-order resonances and calculate the time-averaged fluxes for $50$ different initial phases. We evolve for an integer number of cycles of both the radial and polar motions. Specifically, when considering a $n_\theta$:$n_r$ resonance, we take $T = 2\pi N n_\theta/\Omega_r = 2\pi N n_r/\Omega_\theta$ for some integer $N$. We have found that taking $N=5$ gives sufficiently accurate results. We can then estimate the jumps by taking the difference between the maximum and minimum time-averaged fluxes.

@@

It is informative to consider the dependence on the resonance ratio $\nu$. Fixing the eccentricity at $e=0.95$ to emphasise the variation, we compute the flux enhancements for different resonance ratios, shown in \figref{res-flux-rms-p-vsnu}, where we also highlight ratios arising from different values of $n_\theta$. We observe a clear relationship between the flux enhancement and $\nu$ for all resonances with $n_\theta = 1$ or $2$, independent of which of those is considered. However, negligibly small flux enhancements are observed for systems with $n_\theta = 3$ or higher. This can be explained by the quadrupolar nature of GWs. A resonant trajectory averaged over time (and consequently $\phi$) takes $n_\theta$ (possibly repeated) values of $\theta$ at each radial turning point. Considering rings of matter at these coordinates, the dominant spherical harmonic describing the resultant mass distribution has $l=n_\theta$ (and $m=0$). Since GWs are generated by the quadrupole moment at leading order, the most significant resonant enhancements will arise from systems with $n_\theta = 2$. Due to the presence of higher harmonics, $n_\theta = 1$ systems also contribute strongly; it is perhaps more appropriate to treat $1$:$x$ resonances as \emph{de facto} $2$:$2x$ resonances, explaining the linked behaviour of the flux enhancements for $n_\theta = 1$ and $2$ systems.

@@

The resonant enhancements to the fluxes of $E$, $L_z$ and $Q$ have been calculated previously using Teukolsky-based methods for a handful of systems Flanagan, Hughes and Ruangeri~\cite{Flanagan2012a}. They consider the 1:2, 1:3, 2:3 and 3:4 resonances for EMRIs with $a=0.9$, either $e=0.3$ or $0.7$, and $\theta_\mathrm{inc} = 20\text{\textdegree}$ or $70$\textdegree. We compare our results to their published values (in their tables I--IV) in \figref{res-flux-FHR}. There is some qualitative agreement between the two approaches: more eccentric orbits give rise to larger enhancements; and the 3:4 resonance has a negligibly small enhancement. However, for the most significant 2:3 resonance, we typically predict larger orbital parameter jumps. Our results in \secref{astrophysics}, regarding the impact of resonances on a population of EMRIs, are therefore conservative predictions, as our model over-estimates the resonance effects.

\begin{figure*}[htbp]
\centering
\includegraphics[width=\textwidth]{res_flux_FHR}
\caption{\label{fig:res-flux-FHR}Relative flux enhancements of $E$ (left column), $L_z$ (middle) and $Q$ (right) for resonance ratios corresponding to the 1:3, 1:2, 2:3 and 3:4 resonances. We show results as computed here with a solid red line, and results from Flanagan, Hughes and Ruangeri (FHR14)~\cite{Flanagan2012a} with a dashed blue line. Each row has a different eccentricity and inclination corresponding to tables I--IV in FHR14. All systems have $a=0.9$.}
\end{figure*}

@@

\subsection{Waveform model \& analysis}
\emph{Can probably shorten this section a bit.}

To generate gravitational waveforms, we employ the numerical kludge (NK) method of Babak {\it{et al}}.~\cite{Babak2007}, augmented to include evolution of the positional elements. We first compute inspiral trajectories and then separately (and not necessarily consistently) calculate the GW emission sourced by a compact object moving along that trajectory. This is quicker and easier to calculate than waveforms using Teukolsky-based methods (currently the most accurate prescription available) and yet gives similar results; overlaps between Teukolsky-based and the best NK waveforms are typically 95\% or higher for a variety of orbits~\cite{Babak2007, Berry2013}.

Following the NK method, we first compute the inspiral trajectory of the compact object around a central Kerr BH using the method of osculating elements detailed in \secref{forced-motion}, with the dissipative part of the post-Newtonian self-force from \secref{self-force}. We then map the Boyer--Lindquist coordinates to spherical polar coordinates in Minkowski, facilitating the use of a flat-spacetime waveform generation technique, the standard quadrupole formula~\cite{Misner1973}.

We expect these waveforms to be sufficiently accurate for our purposes, given the inaccuracy of the self-force model being employed. It would, however, be straight-forward to adopt a different waveform generation scheme, for example by including higher-order multipole moments, should advances be made in determining the self-force.

Once we have established a waveform generation procedure, we can define a natural inner product between two waveforms, $s(t)$ and $h(t)$, according to
\begin{equation}
\label{eq:innerprod}
\overlap{s}{h} = 2 \intd{0}{\infty}{\frac{\tilde{s}(f)\tilde{h}^*(f)+\tilde{s}^*(f)\tilde{h}(f)}{S_n(f)}}{f},
\end{equation}
where $\tilde{s}(f)$ represents the Fourier transform of $s(t)$ and similarly for $\tilde{h}(f)$, and $S_n(f)$ is the one-sided noise power spectral density (PSD)~\cite{Moore2014a}. We are interested in the detectability of EMRIs using future space-based GW detectors and so we use a PSD appropriate for describing the eLISA mission concept. We use the analytic approximation of Amaro-Seoane {\it{et al}}.~\cite{Amaro-Seoane2012a}

In practice, the continuous functions $s(t)$ and $h(t)$ are not available, instead replaced by the time series $s(t_i)$ and $h(t_j)$ where $\{t_i\}$ and $\{t_j\}$ are equally-spaced sets of time values, but which do not necessarily have the same length or spacing. To take this into account, we project the $h(t_j)$ onto $\{t_i\}$ using a cubic interpolation scheme and set any values outside the domain of $\{t_j\}$ to zero. A Planck-taper window function~\cite{McKechan2010} is applied to the resulting time series to reduce unwanted spectral leakage in the Fourier transforms. 
%After tapering the projected waveforms, we then calculate overlaps using \eqnref{innerprod} but with discrete Fourier transforms and the integral over frequency replaced with an appropriate sum over Fourier modes.
We maximise every overlap calculation with respect to some time-offset between the models by computing the correlation between the waveforms, which can be performed cheaply using Fourier transforms. %and can provide a significant increase to the overlap, even though typical offset times are shorter than the orbital timescale.

@@

We wish to test whether there exists some set of parameters $\Theta^a$ such that the resulting adiabatic waveform is sufficiently similar to the full waveform. We do this by evaluating the signal-to-noise ratio (SNR) for each waveform template,
\begin{equation}
\rho\left[h\right] = \frac{\overlap{s}{h}}{\sqrt{\overlap{h}{h}}}.
\end{equation}
If a template were to exactly match the data, it would produce an SNR of $\sqrt{\overlap{s}{s}}$ and so the overlap
\begin{equation}
\label{eq:overlap}
\mathbb{O}\left[h\right] = \frac{\overlap{s}{h}}{\sqrt{\overlap{s}{s}}\sqrt{\overlap{h}{h}}}
\end{equation}
provides an indication of how well-matched the template is to the data. We leave an investigation into parameter estimation biases (where the parameters of the best-matching adiabatic $\Theta^a$ do not agree with the parameters of the system $\bar{\Theta}^a$) to future work.

To demonstrate the approach, we can calculate the full evolution and the 2-torus averaged adiabatic evolution over $2$ years for a set of initial parameters chosen to avoid any significant resonances. Example evolutions of the orbital parameters $E$, $L_z$ and $Q$ are shown in \figref{good-traj}, which shows that the adiabatic evolution closely matches the full evolution on longer inspiral timescales. The inset plots show the start of the evolution, on a timescale associated with the orbital motion of the CO; the 2-torus averaging explicitly smooths out the visible structure on this scale.

\begin{figure*}[htbp]
\centering
\includegraphics[width=\textwidth]{Fig_good_traj}
\caption{\label{fig:good-traj}The evolution of the orbital parameters $E$ (left), $L_z$ (center) and $Q$ (right) under the full (solid line) and adiabatic (dashed) models for an illustrative EMRI system that does not encounter any significant resonances. The inset plots show the behaviour on short timescales, where the fast orbital oscillations can be seen. This system has $\mu = 10 M_{\odot}$, $q = 3\times 10^{-6}$, $a_\ast=0.95$, initial semi-latus rectum $p_0=7.5$, initial eccentricity $e_0=0.7$, initial inclination $\cos \iota_0 = 0.5$ and redshift $z=0.204$.}
\end{figure*}

@@

\begin{figure*}[htbp]
\centering
\includegraphics[width=0.92\textwidth]{Fig_good_waveform}
\caption{\label{fig:good-waveform}The plus-polarised waveform for an illustrative EMRI system with initial semi-latus rectum $p_0=7.5$, shown for two short segments at the start and end of a $2$ year evolution. The full and adiabatic models are both shown, but are indistinguishable by eye.}
\end{figure*}

@@

We generate a representative EMRI population by making reasonable assumptions about the distributions of the 14 parameters in the analytic kludge model of Barack and Cutler \cite{Barack2004}. The EMRI event rate is dominated by the inspiral of heavier BHs due to the larger observational volume in comparison to that for white dwarfs and neutron stars~\cite{Gair2004}. We therefore take the mass of the compact object to be $\mu = 10 M_\odot$, corresponding to a typical mass for stellar BHs.

@@

Despite the difficulties, simple estimates of $\mathcal{R}$ have been carried out using Monte Carlo methods to count the number of stars from isothermal distributions that spiral in to a SMBH without plunging. The result is a scaling law for each species of compact object of the form
\begin{equation}
\label{eq:EMRI-intrinsic-rate}
\mathcal{R}(M) = \mathcal{R}_0 \left(\frac{M}{M_0}\right)^\alpha,
\end{equation}
where $M$ is the mass of the SMBH and $M_0 = 3\times 10^6 M_\odot$ is some fiducial mass. \Citet{Hopman2009a} finds that $\alpha = \{-0.15,\,-0.25,\,-0.25\}$ for BHs, NSs and WDs, with respective event rates $\mathcal{R}_0 = \{400,\, 7,\, 20\}\, \mathrm{Gyr}^{-1}$ for each component. This neglects the effects of resonant relaxation, ignores the spin of the SMBH, and assumes that the $M$--$\sigma$ relation holds for all SMBH masses. Each of these is likely to significantly impact the event rate, but \eqnref{EMRI-intrinsic-rate} can still be used as a rough guide to the expected number of events.

\Citet{Amaro-Seoane2011d} studied the effects of mass segregation on the intrinsic EMRI event rate, using direct-summation $N$-body simulations to calibrate a Fokker-Planck description for the bulk properties of the stellar distribution. They found a better fit for the power law spectral index for BHs of $\alpha = -0.19$.

The comoving number density of galaxies is challenging to estimate due to local structure in the Universe, the evolution of that structure, and properties of the SMBHs themselves. We simplify the problem by assuming a homogeneous distribution that does not evolve with redshift, which is reasonable for the typical scales considered by GW detectors. We also neglect correlations between the SMBH mass and spin, and impose a power law scaling relation for the comoving number density
\begin{equation}
\diff{n}{\ln M}(M) = n_0 \left(\frac{M}{M_0}\right)^\beta.
\end{equation}
There is significant uncertainty in the SMBH mass function, but this simple functional form is found to be in good agreement with observations from the Sloan Digital Sky Survey for the mass range of interest; it is sufficient to set $\beta = 0$ and $n_0 = 0.002 \, \mathrm{Mpc}^{-3}$ for SMBHs with $M < \mathcal{O}(10^7 M_\odot)$ \citep{Greene2007, Gair2010b}.
